---
title: "MVCAA Tutorial 1: Exploring PurpleAir Data"
author: "Mazama Science"
date: "2023-07-18"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MVCAA Tutorial 1: Exploring PurpleAir Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

## Introduction

This tutorial demonstrates how to create  _pas_, _pat_ and
_monitor_ objects for PurpleAir sensors in a particular community. 
Target audiences include grad students, researchers, air 
quality professionals and any member of the public concerned about air quality 
and comfortable working with **R** and RStudio.

## Goal

Our goal in this tutorial is to create _pas_, _pat_ and _monitor_ objects
(data structures) for the Methow Valley -- a community in north-central 
Washington state.
[Clean Air Methow](https://www.cleanairmethow.org) 
operates as a project of the
[Methow Valley Citizens Council](http://mvcitizens.org)
and began deploying Purple Air Sensors in 2018:

> In the summer of 2018, Clean Air Methow launched the Clean Air Ambassador 
> Program, an exciting citizen science project, and one of the largest, rural 
> networks of low-cost sensors in the world!

This tutorial will demonstrate how to access and work with data from this 
collection of sensors.

## Data Structures

_pas_ objects in the **AirSensor** package contain per-instrument 
metadata for a collection of PurpleAir sensors. This will include both "spatial 
metadata" like longitude, latitude, timezone, _etc._ as well as each instrument's 
`sensor_index`, allowing us to request timeseries data from the PurpleAir API.

_pat_ objects in the **AirSensor** package contain time-invariant 
spatial metadata for a single sensor as well as the time-dependent 
measurements made by that sensor.

_monitor_ objects in the **AirMonitor** package contain time-invariant 
spatial metadata for multiple sensors as well as the time-dependent PM2.5
measurements made by those sensors.

## Create a PurpleAirSynoptic (pas) object

To find the sensors we wish to investigate, we must first create a _pas_ object 
with metadata for all PurpleAir senors in our target area. The Methow Valley 
valley is located entirely within Okanogan County, WA, so we can create an
Okanogan-only _pas_ object as our starting point.

```{r okanogan_pas, warning = FALSE, message = FALSE}
# AirSensor2 package
library(AirSensor2)

# Set user's PurpleAir_API_READ_KEY
source('global_vars.R')

# Initialize spatial datasets
initializeMazamaSpatialUtils()

# Create a new 'pas' object for Okanogan county
okanogan_pas <-
  pas_createNew(
    api_key = PurpleAir_API_READ_KEY,
    countryCodes = "US",
    stateCodes = "WA",
    counties = c("Okanogan"),
    lookbackDays = 365,         # all sensors from the past year
    location_type = 0           # outdoor sensors only
  )

# Interactive map
okanogan_pas %>% pas_leaflet()
```

### Explore available sensors

Clicking on some of the sensors in the Methow Valley, it quickly becomes 
apparent that many of those sensors have a label that associates them with the
"Clean Air Ambassador" program. Unfortunately, the naming is not consistent.

A quick review of `sort(pas$locationName)` reveals:

* "Clean Air Ambassador @..."
* "MV Ambassador @..."
* "MV Ambassadors@..."
* "MV Clean Air Abassador @..."
* "MV Clean Air Ambassador @..."
* "MV Clean Air Ambassador-..."
* "MVCC Ambassador @..."
* "MVCC Ambassador@..."

Clearly, some effort was made to systematize the naming even if it wasn't 
entirely successful. Nevertheless, we can filter for all location names that
begin with "MV" or "Clean Air" to create a MVCAA-only _pas_ object

```{r mvcaa_pas, warning = FALSE, message = FALSE}
mvcaa_pas <-
  okanogan_pas %>%
  pas_filter(stringr::str_detect(locationName, "^MV|^Clean Air"))

# Interactive map
mvcaa_pas %>% pas_leaflet()
```

## Create a PurpleAirTimeseries (pat) object

A _pat_ object contains time series data for a specific sensor. The
`pat_createNew()` function downloads all data records for a sensor -- "raw data"
typically measured every 2 minutes. A similar function, `pat_createHourly()`,
downloads _hourly aggregated_ data as provided by the PurpleAir API.

### Raw Data

The `pat_createNew()` function has a `fields` argument that lets you specify
which data fields should be included in the result. But default, it uses all
those defined in `PurpleAir_HISTORY_PM25_FIELDS`:

```{r PurpleAir_HISTORY_PM25_FIELDS, echo = FALSE}
PurpleAir_HISTORY_PM25_FIELDS %>%
  stringr::str_split_1(',')
```

Clicking on the leaflet map above, we identify the `sensor_index` for the
"Winthrop Library" sensor as `"13681"`. The following chunk of code creates
a raw _pat_ object for this sensor:

```{r create_pat_raw, warning = FALSE, message = FALSE}
# Create raw pat object
pat <-
  pat_createNew(
    api_key = PurpleAir_API_READ_KEY,
    pas = mvcaa_pas,
    sensor_index = "13681",
    startdate = "2023-07-16",
    enddate = "2023-07-18",
    timezone = "UTC",
    verbose = TRUE
  )

# Pull out data
tbl <- pat$data

# Rewview parameters
names(tbl)
```

We can now use the standard behavior of the base `plot()` function to review
all parameters and look for any interesting correlations among them. 

In the plots below, we see that `temperature` and `humidity` (aka "relative humidity")
are inversely correlated, that `pm2.5_atm_a` and `pm2.5_atm_b` are strongly correlated
and that variables `pm2.5_atm` and `pm2.5_cf` are 
essentially identical.

```{r explore_pat_raw, warning = FALSE, message = FALSE}
# NOTE:  Using "pch = 15" greatly improves the speed of drawing

# Sensor Electronics
plot(tbl[,c(1,2:5)], pch = 15, cex = 0.5, main = "Sensor Electronics")

# Atmospheric Variables
plot(tbl[,c(1,6:8)], pch = 15, cex = 0.5, main = "Atmospheric Variables")

# PM2.5 "atm"
plot(tbl[,c(1,12:14)], pch = 15, cex = 0.5, main = "PM2.5 'atm'")

# PM2.5 comparison
plot(tbl[,c(1,9,12,15)], pch = 15, cex = 0.5, main = "PM2.5 'alt' vs 'atm' vs 'cf1'")
```

## Create a 'monitor' object

For use cases involving comparison with regulatory monitors, calculating daily
averages or informing the public, it is imperative to use hourly aggregated
data that has had a correction equation applied. _(PurpleAir sensors tend to
report pm2.5 values that are higher than those reported by EPA regulatory monitors.)_

The `PurpleAir_createMonitor()` function works similarly to `pat_createHourly()`
but only downloads those parameters typically used in QC and correction functions.
A `correction_FUN` is applied to the hourly _pat_ object and returns corrected
PM2.5 values. By default, an EPA vetted correction function is applied that 
brings PurpleAir values in line with EPA data even in the smoky conditions seen
with wildfire smoke. (See the documentation for `PurpleAir_correction()` for
details.)

`PurpleAir_createMonitor()` returns a _monitor_ object ready to use with the
[AirMonitor](https://github.com/MazamaScience/AirMonitor) and
[AirMonitorPlots](https://github.com/MazamaScience/AirMonitorPlots) packages.

Below, we create a _monitor_ object for a period in August, 2021 when wildfire 
smoke severely impacted the Methow Valley.

_NOTE:  Many sensors will not have data going back multiple years._

```{r create_monitor, warning = FALSE, message = FALSE}
# "Mazama Trailhead" during the Cedar Creek and Cub Creek fires of 2021
monitor <-
  PurpleAir_createNewMonitor(
    api_key = PurpleAir_API_READ_KEY,
    pas = mvcaa_pas,
    sensor_index = "95227",           # Mazama Trailhead
    startdate = "2021-07-15",
    enddate = "2021-08-01",
    timezone = "America/Los_Angeles", # timestamps interpreted in this time zone
    verbose = TRUE
  )

# Check to see that we have data
dim(monitor$data)
```
We can now use functions from the **AirMonitor** and **AirMonitorPlots**
packages to manipulate and visualize this data.

_NOTE: Values are in units of &micro;g/m&sup3; not AQI._

```{r monitor_analysis, warning = FALSE, message = FALSE}
# Create a basic timeseries plot
AirMonitor::monitor_timeseriesPlot(
  monitor,
  shadedNight = TRUE, 
  addAQI = TRUE
)
AirMonitor::addAQILegend(cex = 0.8)

# Create a daily barplot
AirMonitor::monitor_dailyBarplot(
  monitor,
)
AirMonitor::addAQILegend("topleft", cex = 0.8)

# Extract regulatory daily averages using "Local Standard Time"
monitor %>%
  AirMonitor::monitor_dailyStatistic(
    FUN = mean,
    minHours = 18,
    dayBoundary = "LST"
  ) %>%
  AirMonitor::monitor_getData()

# Use AirMonitorPlots to create a "diurnal" plot
monitor %>%
  AirMonitorPlots::monitor_ggDailyByHour_archival()
```

-----

_Best of luck assessing air quality in your community!_
